### 2021-09-08 thesis/tex/oscillations.tex
update

We construct the PMNS matrix using the rotation matrixes from $SO(3)$ as
\begin{align}
    U &= R_{23}R_{13,\delta_{\text{CP}}}R_{12} \nonumber \\
      & = 
    \begin{pmatrix}1 & 0 & 0 \\ 0 & c_{23} & s_{23} \\ 0 & -s_{23} & c_{23}\end{pmatrix}
\begin{pmatrix}c_{13} & 0 & s_{13} e^{-i \delta_{\mathrm{CP}}} \\ 0 & 1 & 0 \\ -s_{13} e^{i \delta_{\mathrm{CP}}} & 0 & c_{13}\end{pmatrix}
\begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1\end{pmatrix}\,.
\end{align}
The term $\delta_\text{CP}$ determines the degree of charge-parity violation. In this work, we will assume CP invariance by always
setting $\delta_{CP} = \SI{0}{\degree}$. Thus,

@@

The solution to Eq.~\ref{eq:TDSE} gives ut the time evolution
in the form of plane wave solutions: 
\begin{align}
    \ket{\nu_k(t)} = e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Inserting the plane wave solution into Eq.~\ref{eq:osc_1}, we get 
\begin{align}\label{eq:nat}
    \ket{\na(t)} = \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Now we know how to evolve and combine the mass eigenstates to form a flavor eigenstate, but how about the reverse?
 We swap the index $k\to j$ in Eq.~\ref{eq:osc_1} and multiply by $U_{\a k}$:
\begin{align}\label{eq:osc_k}
    \sum_\a U_{\a k}\ket{\na} &= \sum_{\alpha, j} U_{\a k} U_{\a j}^\dagger \ket{\nu_j} \nonumber \\
                              &= \sum_{j} \delta_{kj} \ket{\nu_j} \nonumber \\
                              &= \ket{\nu_k}\,,
\end{align}
where we have used the unitarity of the leptonic mixing matrix. Eqs.~\ref{eq:osc_1} and \ref{eq:nat} yield 
\begin{align}
    \ket{\na(t)} &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k} \nonumber \\
                 &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\left(\sum_\b U_{\b k}\ket{\nb}\right) \nonumber \\
                 &= \sum_{k,\b} U^\dagger_{\a k}U_{\b k} e^{-iE_kt} \ket{\nb}\,.
\end{align}
The probability of the flavor transition $\na \to \nb$ at time $t$ is $\abs{\bra{\nb}\ket{\na (t)}}^2$:
\begin{align}
    P_{\na \to \nb}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^*_{\b j}U_{\a j} e^{-i(E_k-E_j)t}\,.
\end{align}

@@

>>>>

We construct the PMNS matrix using the rotation matrixes from $SO(3)$ as
\begin{align}\label{PMNS_def}
    U &= R_{23}R_{13,\delta_{\text{CP}}}R_{12} \nonumber \\
      & = 
    \begin{pmatrix}1 & 0 & 0 \\ 0 & c_{23} & s_{23} \\ 0 & -s_{23} & c_{23}\end{pmatrix}
\begin{pmatrix}c_{13} & 0 & s_{13} e^{-i \delta_{\mathrm{CP}}} \\ 0 & 1 & 0 \\ -s_{13} e^{i \delta_{\mathrm{CP}}} & 0 & c_{13}\end{pmatrix}
\begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1\end{pmatrix}\,.
\end{align}
The term $\delta_\text{CP}$ determines the degree of charge-parity violation. In this work, we will assume CP invariance by always
setting $\delta_{CP} = \SI{0}{\degree}$. Thus,

@@

The solution to Eq.~\ref{eq:TDSE} gives ut the time evolution
in the form of plane wave solutions: 
\begin{align}
    \ket{\nu_k(t)} = e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Inserting the plane wave solution into Eq.~\ref{eq:osc_1}, we get 
\begin{align}\label{eq:nat}
    \ket{\na(t)} = \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Now we know how to evolve and combine the mass eigenstates to form a flavor eigenstate, but how about the reverse?
 We swap the index $k\to j$ in Eq.~\ref{eq:osc_1} and multiply by $U_{\a k}$:
\begin{align}\label{eq:osc_k}
    \sum_\a U_{\a k}\ket{\na} &= \sum_{\alpha, j} U_{\a k} U_{\a j}^\dagger \ket{\nu_j} \nonumber \\
                              &= \sum_{j} \delta_{kj} \ket{\nu_j} \nonumber \\
                              &= \ket{\nu_k}\,,
\end{align}
where we have used the unitarity of the leptonic mixing matrix. Eqs.~\ref{eq:osc_1} and \ref{eq:nat} yield 
\begin{align}
    \ket{\na(t)} &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k} \nonumber \\
                 &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\left(\sum_\b U_{\b k}\ket{\nb}\right) \nonumber \\
                 &= \sum_{k,\b} U^\dagger_{\a k}U_{\b k} e^{-iE_kt} \ket{\nb}\,.
\end{align}
The probability of the flavor transition $\na \to \nb$ at time $t$ is $\abs{\bra{\nb}\ket{\na (t)}}^2$:
\begin{align}
    P_{\na \to \nb}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(E_k-E_j)t}\,.
\end{align}

@@

\subsection{Two neutrino mixing}
By working in the limit where one of the neutrino masses can be neglected, 
we can reduce the complexity of the analytical solution of neutrino oscillation. We now only consider the mixing between two neutrino flavors, say between $\ne, \nm$, and $\nu_1, \nu_2$. In other words, we let $\dm[31] = 0$.

The three-neutrino oscillation Hamiltonian from Eq.~\eqref{eq:H_3gen} %TODO: derive this
now reduces to 
\begin{align}
    H_F &= \frac{1}{2E}(U M^2 U^\dagger + A) \nonumber \\
        &= \frac{1}{2E}\left[U \begin{pmatrix}
            0 & 0 & 0 \\
            0 & \dm_{21} & 0 \\
            0 & 0 & 0
        \end{pmatrix} U^\dagger\right] + \sqrt{2}G_F N_e \begin{pmatrix}
            1 & 0 & 0 \\
            0 & 0 & 0 \\
            0 & 0 & 0
        \end{pmatrix}\,. 
\end{align}
The rotation matrices $R_{ij}$ that constitute $U$ according to Eq.~\eqref{PMNS_def} can be simplified. The rotation $R_{13}$ will not affect our mass matrix, since the former acts in the $1-3$ plane, and the latter sits in the $1-2$ plane. Thus, the reduced mass matrix $\mathcal{M}_{12}$ commutes with $R_{13}$. Analogously, the  rotation matrix $R_{23}$ also commutes with $\mathcal{M}_{12}$.
Moreover, since $SO(3)$ is abelian, its generators commute, i.e.
\begin{align}
    \commutator{R_{ij}}{R_{kl}} = R_{ij}R_{kl} - R_{kl}R_{ij} = 0\,.
\end{align}
Thus,
\begin{align}
    U \mathcal{M}_{12} U^\dagger &=
    R_{23}R_{13}R_{12} \mathcal{M}_{12}  R_{12}^\dagger R_{13}^\dagger  R_{23}^\dagger \nonumber \\
    &= R_{12}R_{23}R_{13} \mathcal{M}_{12}  R_{13}^\dagger R_{23}^\dagger R_{12}^\dagger   \nonumber \\
    &= R_{12}\mathcal{M}_{12} R_{23} R_{13}   R_{13}^\dagger R_{23}^\dagger R_{12}^\dagger  \,.
\end{align}
And due to the fact that the rotation matrices are unitary,
\begin{align}
    R\,R^\dagger = \mathcal{I}\,,
\end{align}
we have that 
\begin{align}
    U \mathcal{M}_{12} U^\dagger &= R_{12}\mathcal{M}_{12}R_{23} R_{13}   R_{13}^\dagger R_{12}^\dagger  R_{23}^\dagger \nonumber \\
    &= R_{12}\mathcal{M}_{12}R_{23} R_{23}^\dagger  R_{12}^\dagger \nonumber \\
    &= R_{12}\mathcal{M}_{12}  R_{12}^\dagger\,.
\end{align}
We have now completely removed the dependence of the mixing angle between the $\nu_1$ and $\nu_3$ mass eigenstates, $\theta_{13}$, and the dependece of the mixing angle between the $\nu_2$ and $\nu_3$ mass eigenstates, $\theta_{23}$.
We can see this clearly by writing our reduced mixing matrix, which is only constituted by the rotation in the $1-2$ plane, $R_{12}$.

\begin{align}
    U &= R_{12}
\begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1 \end{pmatrix} 
\end{align}

Let us return to the Hamiltonian with our two-neutrino mixing matrix. We have 
\begin{align}
    H_F
        &= \frac{1}{2E}\left[R_{12} \begin{pmatrix}
            0 & 0 & 0 \\
            0 & \dm_{21} & 0 \\
            0 & 0 & 0
        \end{pmatrix} R_{12}^\dagger\right] + \sqrt{2}G_F N_e \begin{pmatrix}
            1 & 0 & 0 \\
            0 & 0 & 0 \\
            0 & 0 & 0
        \end{pmatrix} \nonumber \\
        &= \frac{1}{2E}\left[\begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1 \end{pmatrix}  \begin{pmatrix}
            0 & 0 & 0 \\
            0 & \dm_{21} & 0 \\
            0 & 0 & 0
        \end{pmatrix} \begin{pmatrix}c_{12} & s_{12} & 0 \\ -s_{12} & c_{12} & 0 \\ 0 & 0 & 1 \end{pmatrix}^\dagger \right] + \sqrt{2}G_F N_e \begin{pmatrix}
            1 & 0 & 0 \\
            0 & 0 & 0 \\
            0 & 0 & 0
        \end{pmatrix}\,. 
\end{align}

We can reduce this down to two dimensions by extracting the upper-left block of all matrices. We now have the final two-neutrino Hamiltonian:

\begin{align}
    H_F
        &= \frac{1}{2E}\left[
            \begin{pmatrix}
                c_{12} & s_{12} \\ 
                -s_{12} & c_{12} \\ 
            \end{pmatrix}  
            \begin{pmatrix}
            0 & 0 \\
            0 & \dm_{21}
        \end{pmatrix} 
        \begin{pmatrix}
            c_{12} & s_{12}\\ 
            -s_{12} & c_{12} \end{pmatrix}^\dagger \right] + \sqrt{2}G_F N_e 
            \begin{pmatrix}
            1 & 0 \\
            0 & 0 \\
        \end{pmatrix} \nonumber \\
        &= 
        \frac{1}{2E}\left[
            \begin{pmatrix}
                0 & \dm_{21} s_{12} \\ 
                0 & \dm_{21} c_{12} \\ 
            \end{pmatrix}  
        \begin{pmatrix}
            c_{12} & -s_{12}\\ 
            s_{12} & c_{12} \end{pmatrix} \right] + \sqrt{2}G_F N_e 
            \begin{pmatrix}
            1 & 0 \\
            0 & 0 \\
        \end{pmatrix} \nonumber \\
        &= 
        \frac{1}{2E}\left[
            \begin{pmatrix}
                s_{12}^2 + \sqrt{2}G_F N_e & \dm_{21} s_{12}c_{12} \\ 
                \dm_{21} s_{12}c_{12} & \dm_{21} c_{12}^2 \\ 
            \end{pmatrix}  \right]
\end{align}
%TODO: complete trig simplification

This matrix has eigenvalues which we identify as the effective two-neutrino squared-mass difference $(\dm[21])^M$. They are

\begin{align}\label{eq:eff_dm}
    \pm (\dm[21])^M = \pm \sqrt{(\dm[21] \cos{(2\theta_{12}) - \sqrt{2}G_F N_e})^2 + (\dm[21]\sin{(2\theta_{12})})^2}
\end{align}

The matrix is diagonalized by the usual rotation matrix $R_{12}$, but with a different rotation angle. This now introduces the effective two-neutrino mixing angle $\theta_{12}^M$:

\begin{align}
    U^M &= R_{12}^M = R_{12}(\theta_{12}^M)\,.
\end{align}
$U^M$ diagonalizes $H_F$ if and only if

\begin{align}\label{eq:eff_theta}
    \tan{(2\theta_{12}^M)} = \frac{\tan{(2\theta_{12})}}{1- \frac{\sqrt{2}G_F N_e}{\dm[21]\cos{(2\theta_{12})}}}\,.
\end{align}

Now assume that the matter density is sufficiently smooth along the trajectory $x$, that is $\dd \theta_M / \dd x \approx 0$.
In its diagonalized form, the Schrödinger equation with two-neutrino Hamiltonian can be solved with this assuption. The solutions are the regular neutrino vacuum oscillations of Eq.~\eqref{eq:Pab} using $\theta[13] = \theta[23] =0$, but with the effective parameters of Eqs.~\eqref{eq:eff_dm,eq:eff_theta}. That is
%TODO:maybe expand eq below or solve above explicitly?
\begin{align}
    P_{e\mu}^M(x) = \sin^2(2\theta_{12}^M))\sin^2(\frac{\dm[21]x}{4E})
\end{align}

\subsection{Earth propagation}
